Final project


Code to check deforestation based on raster calculations and semi-annual NDVI layers

```{r}
library(gdalcubes)
library(magrittr)
gdalcubes_options(threads=8)

IMAGE_DIR = "L8_cropped" # please change

col = create_image_collection(list.files(IMAGE_DIR, recursive = TRUE, pattern=".tif", full.names  = TRUE), "L8_SR")

# only use "clear" pixels
L8.clear_mask = image_mask("PIXEL_QA", values=c(322, 386, 834, 898, 1346, 324, 388, 836, 900, 1348), invert = TRUE)

# example yearly data cube at 25m spatial resolution
v25 = cube_view(srs="EPSG:3857", extent=col, dx=25, dy=25, dt="P6M", resampling = "average", aggregation = "median")

# calculate NDVI and export as GeoTIFF files at subfolder "L8cube"
raster_cube(col, v25, L8.clear_mask) %>%
  select_bands(c("B04", "B05")) %>%
  apply_pixel("(B05-B04)/(B05+B04)") %>%
  write_tif("L8cube/semi_annual/25m",prefix = "NDVI_")
```

We can read this with package stars, and visualise as:

```{r}
# remotes::install_github("r-spatial/stars")
library(stars)
subdir = "L8cube/semi_annual/25m"
f = paste0(subdir, "/", list.files(subdir))
(st = read_stars(f))
plot(merge(st))
plot(merge(st), breaks = "equal")
```

Getting the Forest layer by checking if a pixel got a NDVI higher than a threshold, to then find deforested areas from half-year to half-year.
```{r}
library(stars)
library(raster)
library(ggplot2)
threshold = 0.8
subdir = "/L8cube/25m"
f = paste0(subdir, "/", list.files(subdir))
(st = read_stars(f))
image2013_04 =raster(x = "L8cube/semi_annual/25m/NDVI_2013-04.tif")
image2014_04 =raster(x = "L8cube/semi_annual/25m/NDVI_2014-04.tif")
image2015_04 =raster(x = "L8cube/semi_annual/25m/NDVI_2015-04.tif")
image2016_04 =raster(x = "L8cube/semi_annual/25m/NDVI_2016-04.tif")
image2017_04 =raster(x = "L8cube/semi_annual/25m/NDVI_2017-04.tif")
image2018_04 =raster(x = "L8cube/semi_annual/25m/NDVI_2018-04.tif")
image2019_04 =raster(x = "L8cube/semi_annual/25m/NDVI_2019-04.tif")
image2013_10 =raster(x = "L8cube/semi_annual/25m/NDVI_2013-10.tif")
image2014_10 =raster(x = "L8cube/semi_annual/25m/NDVI_2014-10.tif")
image2015_10 =raster(x = "L8cube/semi_annual/25m/NDVI_2015-10.tif")
image2016_10 =raster(x = "L8cube/semi_annual/25m/NDVI_2016-10.tif")
image2017_10 =raster(x = "L8cube/semi_annual/25m/NDVI_2017-10.tif")
image2018_10 =raster(x = "L8cube/semi_annual/25m/NDVI_2018-10.tif")
image2019_10 =raster(x = "L8cube/semi_annual/25m/NDVI_2019-10.tif")

hist2013_04 <- hist(image2013_04, plot = FALSE)
hist2014_04 <- hist(image2014_04, plot = FALSE)
hist2015_04 <- hist(image2015_04, plot = FALSE)
hist2016_04 <- hist(image2016_04, plot = FALSE)
hist2017_04 <- hist(image2017_04, plot = FALSE)
hist2018_04 <- hist(image2018_04, plot = FALSE)
hist2019_04 <- hist(image2019_04, plot = FALSE)
hist2013_10 <- hist(image2013_10, plot = FALSE)
hist2014_10 <- hist(image2014_10, plot = FALSE)
hist2015_10 <- hist(image2015_10, plot = FALSE)
hist2016_10 <- hist(image2016_10, plot = FALSE)
hist2017_10 <- hist(image2017_10, plot = FALSE)
hist2018_10 <- hist(image2018_10, plot = FALSE)
hist2019_10 <- hist(image2019_10, plot = FALSE)

plot(hist2013_04, col = "green")
plot(hist2014_04, col = "blue")
plot(hist2015_04, col = "red", add = TRUE) 
plot(hist2016_04, col = "purple", add = TRUE) 
plot(hist2017_04, col = "yellow", add = TRUE) 
plot(hist2018_04, col = "orange", add = TRUE) 
plot(hist2019_04, col = "grey", add = TRUE) 
plot(hist2013_10, col = "green")
plot(hist2014_10, col = "blue")
plot(hist2015_10, col = "red", add = TRUE) 
plot(hist2016_10, col = "purple", add = TRUE) 
plot(hist2017_10, col = "yellow", add = TRUE) 
plot(hist2018_10, col = "orange", add = TRUE) 
plot(hist2019_10, col = "grey", add = TRUE) 


forest2013_04 <- image2013_04 >= threshold
forest2014_04 <- image2014_04 >= threshold
forest2015_04 <- image2015_04 >= threshold
forest2016_04 <- image2016_04 >= threshold
forest2017_04 <- image2017_04 >= threshold
forest2018_04 <- image2018_04 >= threshold
forest2019_04 <- image2019_04 >= threshold
forest2013_10 <- image2013_10 >= threshold
forest2014_10 <- image2014_10 >= threshold
forest2015_10 <- image2015_10 >= threshold
forest2016_10 <- image2016_10 >= threshold
forest2017_10 <- image2017_10 >= threshold
forest2018_10 <- image2018_10 >= threshold
forest2019_10 <- image2019_10 >= threshold

numberForest2013_04 <- freq(forest2013_04, digits=0, value=1, useNA='ifany', progress='')
numberForest2014_04 <- freq(forest2014_04, digits=0, value=1, useNA='ifany', progress='')
numberForest2015_04 <- freq(forest2015_04, digits=0, value=1, useNA='ifany', progress='')
numberForest2016_04 <- freq(forest2016_04, digits=0, value=1, useNA='ifany', progress='')
numberForest2017_04 <- freq(forest2017_04, digits=0, value=1, useNA='ifany', progress='')
numberForest2018_04 <- freq(forest2018_04, digits=0, value=1, useNA='ifany', progress='')
numberForest2019_04 <- freq(forest2019_04, digits=0, value=1, useNA='ifany', progress='')
numberForest2013_10 <- freq(forest2013_10, digits=0, value=1, useNA='ifany', progress='')
numberForest2014_10 <- freq(forest2014_10, digits=0, value=1, useNA='ifany', progress='')
numberForest2015_10 <- freq(forest2015_10, digits=0, value=1, useNA='ifany', progress='')
numberForest2016_10 <- freq(forest2016_10, digits=0, value=1, useNA='ifany', progress='')
numberForest2017_10 <- freq(forest2017_10, digits=0, value=1, useNA='ifany', progress='')
numberForest2018_10 <- freq(forest2018_10, digits=0, value=1, useNA='ifany', progress='')
numberForest2019_10 <- freq(forest2019_10, digits=0, value=1, useNA='ifany', progress='')

numberForestPixels <- c(numberForest2013_04,numberForest2014_04,numberForest2015_04,numberForest2016_04,numberForest2017_04,numberForest2018_04,numberForest2019_04,numberForest2013_10,numberForest2014_10,numberForest2015_10,numberForest2016_10,numberForest2017_10,numberForest2018_10,numberForest2019_10)
year <- c("2013 H1", "2013 H2", "2014 H1", "2014 H2", "2015 H1", "2015 H2", "2016 H1", "2016 H2", "2017 H1", "2017 H2", "2018 H1", "2018 H2", "2019 H1", "2019 H2")
data <- data.frame(year,numberForestPixels)
ggplot(data, aes(x=year, y=numberForestPixels),) +
  geom_line()+
  ggtitle("Development of Forest Pixels in the Research Area")


forestStack <- stack(forest2013_04, forest2013_10, forest2014_04, forest2014_10, forest2015_04, forest2015_10, forest2016_04, forest2016_10, forest2017_04, forest2017_10, forest2018_04, forest2018_10, forest2019_04, forest2019_10)

plot(forestStack)

#one pixel has 25m per 25m
pixelDim <- (25*25)
#the total number of pixel could be read from any forestArea. ncell is the number of cells
totalPixel <- ncell(forest2013_04)
#calculate the total area of one raster picture in qm
totalArea <- totalPixel*pixelDim
#qm into ha
totalAreaHa <- totalArea/10000

#calculate the area that is covered by forest in ha in percent 
forestArea2013_04 <- (numberForest2013_04*pixelDim/10000)/totalAreaHa*100
forestArea2014_04 <- (numberForest2014_04*pixelDim/10000)/totalAreaHa*100
forestArea2015_04 <- (numberForest2015_04*pixelDim/10000)/totalAreaHa*100
forestArea2016_04 <- (numberForest2016_04*pixelDim/10000)/totalAreaHa*100
forestArea2017_04 <- (numberForest2017_04*pixelDim/10000)/totalAreaHa*100
forestArea2018_04 <- (numberForest2018_04*pixelDim/10000)/totalAreaHa*100
forestArea2019_04 <- (numberForest2019_04*pixelDim/10000)/totalAreaHa*100
forestArea2013_10 <- (numberForest2013_10*pixelDim/10000)/totalAreaHa*100
forestArea2014_10 <- (numberForest2014_10*pixelDim/10000)/totalAreaHa*100
forestArea2015_10 <- (numberForest2015_10*pixelDim/10000)/totalAreaHa*100
forestArea2016_10 <- (numberForest2016_10*pixelDim/10000)/totalAreaHa*100
forestArea2017_10 <- (numberForest2017_10*pixelDim/10000)/totalAreaHa*100
forestArea2018_10 <- (numberForest2018_10*pixelDim/10000)/totalAreaHa*100
forestArea2019_10 <- (numberForest2019_10*pixelDim/10000)/totalAreaHa*100

forestAreaPercentage <- c(forestArea2013_04,forestArea2013_10,forestArea2014_04,forestArea2014_10,forestArea2015_04,forestArea2015_10,forestArea2016_04, forestArea2016_10,forestArea2017_04,forestArea2017_10,forestArea2018_04,forestArea2018_10,forestArea2019_04,forestArea2019_10)
data <- data.frame(year,forestAreaPercentage)

ggplot(data, aes(x=year, y=forestAreaPercentage),) +
  geom_line()+
  ggtitle("Development of Forest Pixels in the Research (in %)")

 
#function to check if a cell was covered by forest in a "before" raster
# and in "after" raster not covered by forest
# then the function returns 1
# otherwise it returns 0
deforestDetection <- function(before, after) {
  isDeforested <- (before==1 & after == 0)
  return(isDeforested)
}
deforestation2013_10 <- deforestDetection(forest2013_04,forest2013_10)
deforestation2014_04 <- deforestDetection(forest2013_10,forest2014_04)
deforestation2014_10 <- deforestDetection(forest2014_04,forest2014_10)
deforestation2015_04 <- deforestDetection(forest2014_10,forest2015_04)
deforestation2015_10 <- deforestDetection(forest2015_04,forest2015_10)
deforestation2016_04 <- deforestDetection(forest2015_10,forest2016_04)
deforestation2016_10 <- deforestDetection(forest2016_04,forest2016_10)
deforestation2017_04 <- deforestDetection(forest2016_10,forest2017_04)
deforestation2017_10 <- deforestDetection(forest2017_04,forest2017_10)
deforestation2018_04 <- deforestDetection(forest2017_10,forest2018_04)
deforestation2018_10 <- deforestDetection(forest2018_04,forest2018_10)
deforestation2019_04 <- deforestDetection(forest2018_10,forest2019_04)
deforestation2019_10 <- deforestDetection(forest2019_04,forest2019_10)
deforestationStack <- stack(deforestation2013_10, deforestation2014_04, deforestation2014_10,deforestation2015_04, deforestation2015_10,deforestation2016_04, deforestation2016_10,deforestation2017_04, deforestation2017_10,deforestation2018_04, deforestation2018_10,deforestation2019_04, deforestation2019_10)
colors <- c("white","red")

plot(deforestationStack, col=colors, legend=FALSE, main=c("Deforested Areas (red) \n 2013 H1 to 2013 H2","2013 H2 to 2014 H1","2014 H1 to 2014 H2","2014 H2 to 2015 H1","2015 H1 to 2015 H2", "2015 H2 to 2016 H1", "2016 H1 to 2016 H2", "2016 H2 to 2017 H1", "2017 H1 to 2017 H2", "2017 H2 to 2018 H1", "2018 H1 to 2018 H2", "2018 H2 to 2019 H1", "2019 H1 to 2019 H2"))
#mtext("Histograms",side=3,line=-22,outer=TRUE)

deforestationAll <- deforestation2013_10 | deforestation2014_04 |deforestation2014_10 | deforestation2015_04 |deforestation2015_10 | deforestation2016_04 |deforestation2016_10 | deforestation2017_04 |deforestation2017_10 | deforestation2018_04 |deforestation2018_10 | deforestation2019_04 |deforestation2019_10 |

plot(deforestationAll, col=colors, legend=FALSE, main="Overall Devorestion (2013-2019)")
```
Smooth the raster layer with a window function to make the data less brittle.
```{r}
library(raster)
w <- matrix(c(1,1,1,1,1,1,1,1,1), nr=3,nc=3)

functionDeforest <- function (x){
  center <- x[4]
    if (center==0 | is.na(center)) { # handle NA values
    return(center)
  } 

  ncells <- sum(x, na.rm=TRUE)

  if (ncells<5) { #  window with with less than 5 cells die
    return(0)
  } else if (ncells > 4) { # window with 5 or more cells live
    return(1)
  }
}

deforestSmoothing <- function(x) {
	f <- focal(x, w=w, fun=functionDeforest, pad=TRUE, padValue=0)
}

smoothedDeforestation <- deforestSmoothing(deforestationAll)
plot(smoothedDeforestation, col=colors, legend=FALSE, main="Overall Devorestion smoothed (2013-2019)")

wND <- matrix(c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1), nr=5,nc=5)

functionNaturalDeforest <- function (x){
  center <- x[4]
    if (center==0 | is.na(center)) { # handle NA values
    return(center)
  } 

  ncells <- sum(x, na.rm=TRUE)

  if (ncells<20) { #  window with with less than 5 cells die
    return(1)
  } else if (ncells > 19) { # window with 5 or more cells live
    return(2)
  }
}
naturalDeforestingSmoothing <- function(x) {
	f <- focal(x, w=wND, fun=functionNaturalDeforest, pad=TRUE, padValue=0)
}

smoothedNaturalDeforestation <- naturalDeforestingSmoothing(smoothedDeforestation)
colors <- c("white","yellow", "red")
plot(smoothedNaturalDeforestation, legend=FALSE, col=colors, main="Natural Forest Dieback vs. Deforestion (2013-2019)")
```
Now, take a new Satellite Image and process it to check for deforestation
```{r}
library(gdalcubes)
library(magrittr)
library(stars)
library(raster)
library(ggplot2)

imageNEW =raster(x = "L8cube/newImageSAVED/25m/NDVI_2020-05.tif")

#show the histogram of the numbers of pixels with their NDVI value
#histNEW <- hist(imageNEW, plot = FALSE)
#plot(histNEW, col = "green")
threshold <- 0.8
#get a logical layer weather it is forest or not depending on the NDVI threshold
forestNEW <- imageNEW >= threshold
#count the number of cells that are detected as forest
numberForestNEW <- freq(forestNEW, digits=0, value=1, useNA='ifany', progress='')

#Pixeldimension: one pixel has 25m per 25m
pixelDim <- (25*25)
#the total number of pixel could be read from any forestArea. ncell is the number of cells
totalPixel <- ncell(forestNEW)
#calculate the total area of one raster picture in qm
totalArea <- totalPixel*pixelDim
#qm into ha
totalAreaHa <- totalArea/10000

#calculate the area that is covered by forest in ha in percent 
forestAreaNEW <- (numberForestNEW*pixelDim/10000)/totalAreaHa*100

#get the same extend of the layers
forestNEWCroppedSemiannual <- crop(forestNEW, forest2019_04)
forest2019_04Resampled <- resample(forest2019_04,forestNEWCroppedSemiannual, resample='bilinear')

#use the deforestDetection function to calculate the deforestation from the year 2019 to the newly input picture
deforestationNEWSemiannual_04 <- deforestDetection(forest2019_04Resampled, forestNEWCroppedSemiannual)
colors <- c("white","red")
plot(deforestationNEWSemiannual_04, col=colors, legend=FALSE, main="Deforestion in 2020")

smoothedDeforestationNEW_04 <- deforestSmoothing(deforestationNEWSemiannual_04)
smoothedNaturalDeforestationNEW_04 <- naturalDeforestingSmoothing(smoothedDeforestationNEW_04)
colors <- c("white","yellow", "red")
plot(smoothedNaturalDeforestationNEW_04, legend=FALSE, col=colors, main="Natural Forest Dieback vs. Deforestion (2020)")
writeRaster(smoothedNaturalDeforestationNEW_04,'deforestation2020SemiAnnual_04.tif', overwrite=TRUE)
```